---
title: "Gold Mine Analysis - Liam Hash"
output: rmarkdown::github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a file to show how I valued a gold mine, using Monte-Carlo simulations of gold prices and NPV functionality.

### Cleaning the Data

The first part of the process was to load the monthly gold price data for the past 5 years and clean it so that it was in a format that I could use productively in the Monte-Carlo simulations. This also includes loading the necessary libraries for our project.

```{r, echo = TRUE}
library(ggplot2)
library(plotly)
library(fitdistrplus)
```


```{r, echo = FALSE}
# setting working directory to where LBMA-GOLD.csv is stored
setwd("C:/Users/wphny/Desktop/Work/William Hash Portfolio/Portfolio/Gold Mine Valuation")

# reading the gold prices csv into R
gold <- read.csv("gold_prices.csv")

# taking just the date and USD closing prices for gold
gold <- gold[, c(1,2)]

# renaming the columns
names(gold) <- c("Date", "Closing Price")

# changing class of Date from factor to date
gold$Date <- as.Date(gold$Date, format = "%m/%d/%Y")
```

### Understanding the Data

Next, I ran two plots to analyze the past data for gold prices. One shows historical gold prices over the past 5 years, and the other is a histogram analyzing the frequency of gold prices for the past 5 years.

```{r, echo = TRUE}
# creating a chart to show how gold prices have changed over time
ggplot(gold, aes(x = Date, y = `Closing Price`)) + 
            geom_point(color = "gold") + 
            geom_line(color = "gold") +
            ggtitle("Gold Monthly Prices") + 
            xlab("Date") +
            ylab("Close Price ($)") +
            theme_dark()
```

```{r, echo = TRUE}
#creating histogram of gold prices
ggplot(gold) + 
            aes(`Closing Price`) + 
            geom_histogram(binwidth = 25, color = "black", fill = "gold") +
            ylab("Frequency") +
            xlab("Price of Gold") +
            ggtitle("Histogram: Gold Prices") +
            theme_dark()
```

### Fitting the Data

Before I can run Monte-Carlo simulations, I must find the mean and standard deviation of the distribution by looking at the daily percent change in gold prices over the past 5 years.

```
# finding daily % changes in gold prices
gold$return <- rep(0,length(gold$Date))

# creating a column that shows daily percent change in gold prices
for (i in 2:length(gold$Date)){
  gold$return[i] <- (gold$`Closing Price`[i] - gold$`Closing Price`[i-1]) / gold$`Closing Price`[i-1]
}

# using the daily returns vector to characterize the distribution of gold prices
gold_returns_dist <- fitdist(gold$return,"norm", method = c("mle"))
summary(gold_returns_dist)
```
```{r,echo = FALSE}
# finding daily % changes in gold prices
gold$return <- rep(0,length(gold$Date))

# creating a column that shows daily percent change in gold prices
for (i in 2:length(gold$Date)){
  gold$return[i] <- (gold$`Closing Price`[i] - gold$`Closing Price`[i-1]) / gold$`Closing Price`[i-1]
}

# using the daily returns vector to characterize the distribution of gold prices
gold_returns_dist <- fitdist(gold$return,"norm", method = c("mle"))
summary(gold_returns_dist)
```

Now we must take the mean and standard deviation values from the summary table above.
We will use the mean and standard deviation to run the Monte-Carlo simulations.

```{r, echo = TRUE}
# extracting the mean and standard deviation from the summary table
c <- summary(gold_returns_dist)

gold_mean <- c$estimate[1]
gold_sd <- c$estimate[2]
```

### Monte-Carlo Simulations

With the parameters discovered and stored away, now we can use them within the Monte-Carlo simulations.
First we need to define the asset paths function

```{r, echo = TRUE}
# running the Monte-Carlo asset paths function
asset.paths <- function(s0, 
                        mu, 
                        sigma, 
                        nsims = 10000,
                        periods = c(0, 1)) {   # time periods at which to simulate prices
  s0 = as.vector(s0)
  nsteps = length(periods)
  dt = c(periods[1], diff(periods))
  if(length(s0) == 1) {
    drift = mu - 0.5 * sigma^2
    if(nsteps == 1) {
      s0 * exp(drift * dt + sigma * sqrt(dt) * rnorm(nsims))
    } else {
      temp = matrix(exp(drift * dt + sigma * sqrt(dt) * rnorm(nsteps * nsims)), nc = nsims)
      for(i in 2:nsteps) temp[i,] = temp[i,] * temp[(i-1),]
      s0 * temp
    }
  } else {
    require(MASS)
    drift = mu - 0.5 * diag(sigma)
    n = length(mu)
    if( nsteps == 1 ) {
      s0 * exp(drift * dt + sqrt(dt) * t(mvrnorm(nsims, rep(0, n), sigma)))
    } else {
      temp = array(exp(as.vector(drift %*% t(dt)) + t(sqrt(dt) * mvrnorm(nsteps * nsims, rep(0, n), sigma))), c(n, nsteps, nsims))
      for(i in 2:nsteps) temp[,i,] = temp[,i,] * temp[,(i-1),]
      s0 * temp
    }
  }
}
```

We will now define and store values in the necessary variables to run the above function.

```{r, echo = TRUE}
# defining our metrics that will go into the asset price simulations
# use the latest price as s
S = gold$`Closing Price`[62]

# define the mean as mu
mu = gold_mean

# define the sd as sigma
sigma = gold_sd

# run the experiment over 1000 times
N = 1000

# periods will be 240, since it is 20 years * 12 months
periods = (0:240)
```

With these variables, we can run the asset prices function on our data to create our price paths via Monte-Carlo method.

```{r, echo = TRUE}
# running the Monte-Carlo simulation for predicting gold prices
prices = asset.paths(S, mu, sigma, N, periods = periods)
```

Here is the plotted Monte-Carlo price simulations.
```{r, echo = TRUE}
matplot(prices[,1:100], type = 'l', xlab = 'Months', ylab = 'Prices', main='Selected Price Paths for Gold Prices')
```

### Valuing the Gold Mine

To calculate revenues/cash flows for each year, we must decide on a selling price for gold for each year.

```{r, echo = TRUE}
# creating a matrix with prices at yearly intervals
yearly_prices <- prices[, c(12, 24, 36, 48, 60, 72, 84, 96, 108, 120, 132, 144, 156, 168, 180, 192, 204, 216, 228, 240)]

# taking mean values for each column of yearly prices and storing in avg yearly prices vector
avg_yearly_prices <- colMeans(yearly_prices)
avg_yearly_prices
```

To find our profit margin, we must subtract production cost from the average selling price of gold per year.
Once we do this, we can multiply our profit margin per unit by our quantity sold (which is 50,000 per year).

```{r, echo = TRUE}
# creating a vector for production cost per year
# this starts at 250, but rises each year by 5%
production_cost <- rep(0, 20)
for (i in 1:length(production_cost)) {
  production_cost[i] = 250 * ((1.05)^(i-1))
}
production_cost

# creating a vector of profit margin per unit for each year
profit_margin <- rep(0, 20)
for (i in 1:length(profit_margin)) {
  profit_margin[i] = avg_yearly_prices[i] - production_cost[i]
}
profit_margin
```

Now we must create a vector which includes our quantity sold, so we can multiply those values with our profit margin.

```{r, echo = TRUE}
# creating a vector of quantity sold per year
quantity_sold <- rep(50000, 20)
```

We then create a cash flows vector by multiplying profit margin per unit in each year by our quantity sold for each year.

```{r, echo = TRUE}
# creating the cash flows vector
cash_flows <- rep(0, 20)
for (i in 1:length(cash_flows)) {
  cash_flows[i] = profit_margin[i] * quantity_sold[i]
}
cash_flows
```

Now, with our cash flows vector, we must discount them using the NPV method to find the overall value of the future cash flows. We will subtract the initial $100million investment at time 0 to find the overall value of the firm!
For this case, the discount rate we will be using is the 20-year risk free rate, which is 3.02%

```{r, echo = TRUE}
# storing our discount rate variable as r
r = .0302

# discounting the cash flows for each year, via npv
discounted_cf <- rep(0, 20)
for (i in 1:length(discounted_cf)) {
  discounted_cf[i] = cash_flows[i] / ((1 + r)^i)
}
discounted_cf

# taking sum of discounted cash flows to find current value of future cash flows
current_value <- sum(discounted_cf)
current_value

# subtracting the initial $100million investment
value = current_value - 100000000
value
```

#### So, our value for the 20-year gold mine is the above value! This is a positive NPV, so we should go ahead with building the mine.

Now we will check to see the value of the gold mine if we stop the production and close it at year 10.
To do this, we will take the first 10 years of cash flows and discount by the 10-year risk free rate, which is 2.86% and then sum the discounted cash flows, similar to the process above. Once we subtract the initial investment of $100million, we will have the full value.

```{r, echo = TRUE}
# FOR 10-YEAR CLOSING
# storing our discount rate variable as r
r10 = .0286

# discounting the cash flows for each year, via npv
discounted_cf_10 <- rep(0, 10)
for (i in 1:length(discounted_cf_10)) {
  discounted_cf_10[i] = cash_flows[i] / ((1 + r10)^i)
}
discounted_cf_10

# taking sum of discounted cash flows to find current value of future cash flows
current_value_10 <- sum(discounted_cf_10)
current_value_10

# subtracting the initial $100million investment
value_10 = current_value_10 - 100000000
value_10
```

#### So, our value for the 10-year gold mine is the above value! This is a positive NPV, so we should go ahead with building the mine.

Now we want to check the value of the gold mine if we close it at any year. To do this, we will need to create a for loop that sums the discounted cash flows for years 1:i. Although the most accurate way to do this would be to use a different risk-free rate for closing at each year, we decided to use the 10-year risk free rate (2.86%) for all of them, just for simplicity. Then we will follow the same npv process as the last two valuations.

```{r, echo = TRUE}
# FOR ANY-YEAR CLOSING
# storing our discount rate variable as r
rn = .0286

# discounting the cash flows for each year, via npv
discounted_cf_n <- rep(0, 20)
for (i in 1:length(discounted_cf_n)) {
  discounted_cf_n[i] = cash_flows[i] / ((1 + rn)^i)
}
discounted_cf_n

# taking sum of discounted cash flows to find current value of future cash flows
# also subtracting each value by the $100million investment
value_n <- rep(0, 20)
for (i in 1:length(value_n)) {
  value_n[i] = sum(discounted_cf_n[1:i]) - 100000000
}
value_n
```

#### So the above vector shows the values of the gold mine if we close at each year! As it shows, our value continues to grow year on year, with the lowest values coming if we closed the mine immediately.






